---
title: 'Week 4: Activity 1'
author: "Calum Smith"
date: "2026-02-02"
output: html_document
#params:
  #data_version: "updated"
  #group_var: "group"
---

```{r, warning = FALSE, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning = FALSE, message = FALSE}
library(ggplot2)
library(gtsummary)
library(dplyr)
library(gt)
library(janitor)
library(here)
library(readr)
library(tidyverse)
```

# Introduction
This report provides an initial, plain-language assessment of whether your dataset supports a clear and defensible message that could be communicated to a general audience. At this stage we focus on describing what the data show and how confident we can be in that description, rather than applying complex models. In particular, we (i) summarise who is in the dataset, (ii) compare the two groups on the information available, and (iii) highlight any issues that could make a public-facing narrative misleading without further checks.

# Client Brief
You asked us to assess whether the dataset supports a clear message that you can stand behind publicly. In practical terms, you want to know whether the pattern is relevant to your decisions, consistent enough to tell a clear story, and not easily explained by straightforward problems such as the two groups being very different people, missing information, or changes in how the outcome was defined or recorded.

# Data Description


# Baseline summary of data characteristics

# Claim

# Data and approach(plain language)

# Evidence


```{r}
# create file path to data selected
#data_version <- params$data_version
data_version <-  "updated"

data_path <- here("week_4", "data", "raw", paste0("week4_dataset_",
                                                  data_version, ".csv"))

load_dynamic_data <- function(file_path) {
  
  # Read the first line
  first_line <- readLines(file_path, n = 1)
  
  # Count Semicolons vs Commas
  cnt_semi <- length(gregexpr(";", first_line)[[1]])
  cnt_comma <- length(gregexpr(",", first_line)[[1]])
  
  # Choose the correct Tidyverse function
  if (cnt_semi > cnt_comma) {
    message("Detected Semicolon separator. Using read_delim(delim = ';')...")
    df <- read_delim(file_path, delim = ";", show_col_types = FALSE)
  } else {
    message("Detected Comma separator. Using read_csv()...")
    df <- read_csv(file_path, show_col_types = FALSE)
  }
  
  return(df)
}

df <- load_dynamic_data(data_path)

```

```{r}
# Next we need to clean the data across the three sources
df_clean <- df %>%
  mutate(across(where(is.character), as.factor)) %>%  # Ensure strings are factors
  janitor::clean_names()
```

```{r}
# A function to turn "tenure_months" into "Tenure Months"
correct_labels <- function(x) {
  x %>%
    stringr::str_replace_all("_", " ") %>%
    stringr::str_to_title() %>%
    stringr::str_replace("Gbp", "(£)") %>%
    stringr::str_replace("90d", "(90 Days)") %>%
    stringr::str_replace("3m", "(90 Days)")
}

# Apply this to all current column names in the dataset
dynamic_labels <- setNames(
  as.list(correct_labels(names(df_clean))), 
  names(df_clean)
)
```

```{r}
# Column variables
#group_var <- params$group_var
group_var <- "group"

# Building the table 1
table1 <- df_clean %>%
  select(
    all_of(group_var),
    where(is.numeric),
    where(is.factor),
    -any_of(c("customer_id", "signup_date"))
  ) %>%
  tbl_summary(
    by = all_of(group_var),
    label = dynamic_labels,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    missing_text = "(Missing)",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_overall(last = FALSE) %>%
  add_n() %>%
  add_stat_label(location = "row") %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption(paste("**Table 1. Characteristics by", 
                       tools::toTitleCase(group_var), 
                       "(Data Version:", data_version, ")**"))

# Output the Table 1
table1_formatted <- table1 %>% 
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(9),
    table.width = gt::pct(100)
  )

# Save the table
gt::gtsave(
  data = table1_formatted,
  filename = here::here("week_4", "outputs", "tables",
                          paste0("table_1_", data_version, ".html"))
  )

table1_formatted
```

```{r}
# target_var <- params$target_var
target_var <- "avg_order_value_gbp"

boxplot_p <- ggplot(df_clean, aes(x = .data[[group_var]],
                                  y = .data[[target_var]],
                                  fill = .data[[group_var]])) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = paste("Distribution of", dynamic_labels[[target_var]], "by",
                  dynamic_labels[[group_var]]),
    subtitle = paste("Dataset Version:", data_version),
    x = dynamic_labels[[group_var]],
    y = dynamic_labels[[target_var]],
    fill = dynamic_labels[[group_var]]
  ) +
  theme(legend.position = "none")

boxplot_p

ggsave(here::here("week_4", "outputs", "figures",
                  paste0("boxplot_", data_version, ".png")), boxplot_p)
```

```{r}
hist_var <- "web_sessions_last_30d"
dynamic_bins <- ceiling(sqrt(nrow(df_clean)))

histogram_p <- ggplot(df_clean, aes(x = .data[[hist_var]])) +
  geom_histogram(
    fill = "steelblue",
    color = "white",
    bins = dynamic_bins
  ) +
  theme_minimal() +
  labs(
    title = paste("Frequency Distribution of", dynamic_labels[[hist_var]]),
    subtitle = paste0("Dataset: ", data_version, 
                      " (Bins calculated via Square-root rule: ",
                      dynamic_bins, ")"),
    x = dynamic_labels[[hist_var]],
    y = "Frequency"
  )

histogram_p

ggsave(here::here("week_4", "outputs", "figures", 
                  paste0("histogram_",data_version, ".png")), histogram_p)
```
