---
title: 'Week 4: Activity 1'
author: "Calum Smith"
date: "2026-02-02"
output: html_document
params:
  data_version: "updated"
  group_var: "group"
  target_var: "avg_order_value_gbp"
---

```{r, warning = FALSE, message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning = FALSE, message = FALSE}
library(ggplot2)
library(gtsummary)
library(dplyr)
library(gt)
library(janitor)
library(here)
library(readr)
library(tidyverse)
library(smd)
```

# Introduction
This report presents an initial descriptive analysis to assess whether the dataset supports a clear and defensible narrative relevant to the client’s stated objective. The emphasis at this stage is on establishing a transparent evidential basis for subsequent analysis, rather than pursuing complex modelling. Specifically, the report aims to (i) characterise the sample, (ii) assess baseline comparability between groups on observed covariates, and (iii) identify issues that may affect interpretability, including limitations in measurement or coverage.

# Client Brief
The client has requested an assessment of whether the dataset contains a pattern that is relevant to decision-making, sufficiently consistent to justify communication to a broad audience, and not readily explained by straightforward artefacts (for example, changes in who is included, missingness, or definitional changes). The required output is an evidence-led summary with an explicit claim, supporting evidence, limitations with impact, and a concrete recommendation for next steps.

# Data Description
The dataset contains 238 customers split into two groups: Control (n=120) and Treatment (n=118). For each customer, the dataset includes customer age, customer type, average order value in GBP, number of orders placed in the past 90 days, 
number of website sessions in the last 30 days prior to the extract date, whether the customer is eligible for a discount under the intervention rules,  area-level affluence score associated with the customer’s location and whether the customer made at least one purchase in the 30 days after the intervention/extract reference point. The operational meaning of the outcome is assumed to be consistent across records but should be verified against the source system definition.

# Baseline summary of data characteristics
We summarise baseline characteristics by group to assess whether the groups appear comparable on observed covariates. For continuous variables we report means and standard deviations (and median and IQR for deprivation score), and for categorical variables we report counts and percentages. For the expert audience we additionally report standardised mean differences (SMDs) to quantify imbalance; these provide a scale-free measure of the magnitude of observed differences.

# Claim
The dataset supports a valid comparison between groups with no evidence of sampling bias; however, the marginal increase in Average Order Value for the Treatment group is heavily obscured by high customer variability, requiring statistical validation before confirming success.

# Data and approach (plain language)
The analysis is descriptive. We do not fit outcome models or claim causal effects. We use the baseline summary table to characterise the sample and check for obvious differences in group composition, then use two simple visuals: (i) a plot of the customer type across the groups; and (ii) a plot of Average Order Value (£) distributions by group as an example check to illustrate the skewness of the data.


```{r, warning = FALSE, message = FALSE}
# create file suffix to map to the correct file
file_suffix <- if (params$data_version == "original") {
  "" 
} else {
  paste0("_", params$data_version)
}

# choose the file based on the suffix
data_path <- here("week_4", "data", "raw", paste0("week4_dataset",
                                                  file_suffix, ".csv"))

# bespoke data loader to manage different separators in the csv
load_dynamic_data <- function(file_path) {
  
  # Read the first line
  first_line <- readLines(file_path, n = 1)
  
  # Count Semicolons vs Commas
  cnt_semi <- length(gregexpr(";", first_line)[[1]])
  cnt_comma <- length(gregexpr(",", first_line)[[1]])
  
  # Choose the correct Tidyverse function
  if (cnt_semi > cnt_comma) {
    message("Detected Semicolon separator. Using read_delim(delim = ';')...")
    df <- read_delim(file_path, delim = ";", show_col_types = FALSE)
  } else {
    message("Detected Comma separator. Using read_csv()...")
    df <- read_csv(file_path, show_col_types = FALSE)
  }
  
  return(df)
}

# create a data frame
df <- load_dynamic_data(data_path)

```

```{r, warning = FALSE, message = FALSE}
# Next we need to clean the data across the three sources
df_clean <- df %>%
  mutate(across(where(is.character), as.factor)) %>%  # Ensure strings are factors
  janitor::clean_names()
```

```{r, warning = FALSE, message = FALSE}
# A function to turn "tenure_months" into "Tenure Months"
correct_labels <- function(x) {
  x %>%
    stringr::str_replace_all("_", " ") %>%
    stringr::str_to_title() %>%
    stringr::str_replace("Gbp", "(£)") %>%
    stringr::str_replace("90d", "(90 Days)") %>%
    stringr::str_replace("3m", "(90 Days)")
}

# Apply this to all current column names in the dataset
dynamic_labels <- setNames(
  as.list(correct_labels(names(df_clean))), 
  names(df_clean)
)
```

```{r, warning = FALSE, message = FALSE}
# Column variables
group_var <- params$group_var


# Building the table 1
table1 <- df_clean %>%
  select(
    all_of(group_var),
    where(is.numeric),
    where(is.factor),
    -any_of(c("customer_id", "signup_date"))
  ) %>%
  tbl_summary(
    by = all_of(group_var),
    label = dynamic_labels,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    missing_text = "(Missing)",
    missing_stat = "{N_miss} ({p_miss}%)"
  ) %>%
  add_overall(last = FALSE) %>%
  add_n() %>%
  add_difference(
    test = everything() ~ "smd",
    estimate_fun = everything() ~ function(x) style_sigfig(x, digits=3)
  )%>%
  add_stat_label(location = "row") %>%
  modify_header(label ~ "**Variable**",
                estimate = "**SMD**") %>%
  modify_column_hide(columns = c("conf.low", "conf.high")) %>%
  modify_caption(paste("**Table 1. Characteristics by", 
                       tools::toTitleCase(group_var), 
                       "(Data Version:", params$data_version, ")**"))

# Remove CI abbreviation
table1$table_styling$abbreviation <- table1$table_styling$abbreviation %>%
  filter(!grepl("CI|Confidence Interval", abbreviation))

# Output the Table 1
table1_formatted <- table1 %>% 
  as_gt() %>%
  gt::tab_options(
    table.font.size = gt::px(9),
    table.width = gt::pct(100)
  )

# Save the table
gt::gtsave(
  data = table1_formatted,
  filename = here::here("week_4", "outputs", "tables",
                          paste0("table_1_", params$data_version, ".html"))
  )

table1_formatted
```


# Evidence

## Baseline summary table & Interpretation
The baseline summary table reports key covariates by group. For this expert version we include standardised mean differences (SMDs) to quantify baseline imbalance.

**Interpretation**: The SMDs are almost all <0.1 with only Customer Type reaching 0.152. Recorded covariates exhibit negligible to small between-group differences overall; balance is assessed on observed variables only and does not address unmeasured determinants.

# Visualisation 1 & Interpretation

```{r}
ggplot(df_clean, aes(x=customer_type, fill=group)) +
  geom_bar(position="dodge") +
  labs(title = "Customer Types by Group", x = "Type",
       subtitle = paste("Dataset Version:", params$data_version)) +
  theme_minimal()
```

Visualisation 1: Count of Existing and New customers per group.

**Interpretation**: Both groups exhibit similar counts of both Existing and New customers.

# Visualisation 2 & Interpretation

Visualisation 2 compares the Average Order Value (£) distributions by group.
```{r, warning = FALSE, message = FALSE}
target_var <- params$target_var

boxplot_p <- ggplot(df_clean, aes(x = .data[[group_var]],
                                  y = .data[[target_var]],
                                  fill = .data[[group_var]])) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(
    title = paste("Distribution of", dynamic_labels[[target_var]], "by",
                  dynamic_labels[[group_var]]),
    subtitle = paste("Dataset Version:", params$data_version),
    x = dynamic_labels[[group_var]],
    y = dynamic_labels[[target_var]],
    fill = dynamic_labels[[group_var]]
  ) +
  theme(legend.position = "none")

boxplot_p

ggsave(here::here("week_4", "outputs", "figures",
                  paste0("boxplot_", params$data_version, ".png")), boxplot_p)
```

Visualisation 2: Average Order Value (£) by Group.

**Interpretation**: Average Order Value (£) distributions significantly overlap across the groups however, both contain outliers indicating the data for each group is right-skewed. 

# Uncertainty & Limitations
* Influence of Outliers: The AOV distribution is right-skewed with significant outliers. These extreme values can disproportionately pull the mean, potentially masking the typical customer behavior represented by the median.
* Ambiguity in Assignment Rules: The specific logic used to assign customers to "Treatment" or "Control" groups is currently undocumented; if this allocation was based on any non-random criteria, the observed differences in outcomes may be driven by selection bias rather than the intervention itself.

# Recommendation & Next Steps
Sensitivity analysis on outliers to ensure the findings are robust and not driven by the top 1% of spenders. If definitions are stable, proceed to an adjusted analysis aligned to the client decision context before communicating a public-facing narrative.